---
title: "PracCompFinalProject_Sioson"
author: "Michael Sioson"
date: "2024-12-05"
output: html_document
---

# Graphical Abstract
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(dplyr)
#install.packages("MuMIn")
library(MuMIn)
```

```{r}
thesisdesign <- read.csv("thesis_design_pre.csv")
thesisdesign$Addition <- factor(thesisdesign$Addition, 
                                levels = c("No_Addition", "Sulfur_Oxidizers", "Community_Addition"))
ggplot(thesisdesign, aes(x = Inundation, y = Plant_Biomass, fill = Soil)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Addition) +
  labs(
    title = "Expected S. alternflora Biomass by Treatment",
    x = "Inundation",
    y = "Plant Biomass",
    fill = "Soil"
  ) +
  scale_y_continuous(
    breaks = range(thesisdesign$Plant_Biomass),  
    labels = c("Low", "High")
  ) +
  scale_fill_manual(values = c("Organic_Soil" = "chartreuse4", 
                               "Sandy_Soil " = "lightgoldenrod1")) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Spartina Data

## Stem Height

```{r}
set.seed(29)
spartinatest <- read.csv("Spartina_TestData.csv")
spartinatest$Stem_Height_mm <- rnorm(120, mean = 370, sd = 158)

ht_data <- spartinatest %>%
  arrange(Stem_Height_mm) %>%
  mutate(Group = ntile(Stem_Height_mm, 12)) %>%
  
  # Assign Inundated and Soil_Treatment
  mutate(
    Inundated = case_when(
      Group %in% 1:6 ~ "Yes",
      Group %in% 7:12 ~ "No"
    ),
    Soil_Treatment = case_when(
      Group %in% c(1, 2, 3, 7, 8, 9) ~ "Organic",
      Group %in% c(4, 5, 6, 10, 11, 12) ~ "Sand"
    )
  ) %>%
  
  # Assign exactly 10 of each Microbe_Treatment per group
  group_by(Group) %>%
  mutate(
    Microbe_Treatment = case_when(
      Group %in% c(1, 4, 7, 10) ~ "None", # Groups with fixed "None"
      Group %in% c(2, 3) ~ rep(c("SOB", "Community"), times = 10)[row_number()],
      Group %in% c(5, 8, 11) ~ "SOB",
      Group %in% c(6, 9, 12) ~ "Community"
    )
  ) %>%
  ungroup()


ht_data <- ht_data %>% mutate(Stem_Height_mm = if_else(Stem_Height_mm < 105, NA, Stem_Height_mm)) 
```

```{r}
ht_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  summarize(count = n(),
            mean = mean(Stem_Height_mm),
            sd = sd(Stem_Height_mm))

ht_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  mutate(Microbe_Treatment = factor(Microbe_Treatment, levels = c("None", "SOB", "Community"))) %>%
  mutate(Inundated = factor(Inundated, levels = c("Yes", "No"))) %>%
  
ggplot(aes(x = Inundated, y = Stem_Height_mm, fill = Soil_Treatment)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_color_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  scale_fill_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  facet_wrap(~ Microbe_Treatment) +
  labs(
    title = "Stem Height by Microbe Treatment and Inundation",
    x = "Inundation",
    y = "Stem Height (mm)",
    fill = "Soil Treatment"
  ) +
  theme_light()

```

## Belowground Biomass

```{r}
set.seed(29)
bg_data <- rnorm(120, mean = 0.7104, sd = 0.4) %>% sort()  # Generate and sort values

bg_data <- ht_data %>%
  mutate(
    Belowground_biomass_g = {
      group_order <- order(Group)  # Order plants by their Group
      bg_data[group_order]         # Assign sorted biomass to plants by Group rank
    },

    Belowground_biomass_g = if_else(Belowground_biomass_g < 0.04, NA, Belowground_biomass_g)
  )
```

```{r}
bg_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  summarize(count = n(),
            mean = mean(Belowground_biomass_g),
            sd = sd(Belowground_biomass_g))

bg_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  mutate(Microbe_Treatment = factor(Microbe_Treatment, levels = c("None", "SOB", "Community"))) %>%
  mutate(Inundated = factor(Inundated, levels = c("Yes", "No"))) %>% 
  
ggplot(aes(x = Inundated, y = Belowground_biomass_g, fill = Soil_Treatment)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_color_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  scale_fill_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  facet_wrap(~ Microbe_Treatment) +
  labs(
    title = "Belowground Biomass by Microbe Treatment and Inundation",
    x = "Inundation",
    y = "Belowground Biomass (g)",
    fill = "Soil Treatment"
  ) +
  theme_light()
```

## Aboveground Biomass

```{r}
set.seed(29)
ag_data <- rnorm(120, mean = 2.146, sd = 1.3) %>% sort()
ag_data <- bg_data %>%
  
  mutate(
    Aboveground_biomass_g = {
      group_order <- order(Group)  
      ag_data[group_order]         
    },

    Aboveground_biomass_g = if_else(Aboveground_biomass_g < 0, NA, Aboveground_biomass_g)
  )
```
```

```{r}
ag_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  summarize(count = n(),
            mean = mean(Aboveground_biomass_g),
            sd = sd(Aboveground_biomass_g))

ag_data %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  mutate(Microbe_Treatment = factor(Microbe_Treatment, levels = c("None", "SOB", "Community"))) %>%
  mutate(Inundated = factor(Inundated, levels = c("Yes", "No"))) %>%
  
ggplot(aes(x = Inundated, y = Aboveground_biomass_g, fill = Soil_Treatment)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_color_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  scale_fill_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  facet_wrap(~ Microbe_Treatment) +
  labs(
    title = "Aboveground Biomass by Microbe Treatment and Inundation",
    x = "Inundation",
    y = "Aboveground Biomass (g)",
    fill = "Soil Treatment"
  ) +
  theme_light()
```

## Total Biomass

```{r}
sp_data_final <- ag_data %>%
  mutate(Total_biomass_g = Belowground_biomass_g + Aboveground_biomass_g)
```

```{r}
sp_data_final %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  summarize(count = n(),
            mean = mean(Total_biomass_g),
            sd = sd(Total_biomass_g))

sp_data_final %>%
  filter(!is.na(Stem_Height_mm)) %>%
  group_by(Inundated, Soil_Treatment, Microbe_Treatment) %>%
  mutate(Microbe_Treatment = factor(Microbe_Treatment, levels = c("None", "SOB", "Community"))) %>%
  mutate(Inundated = factor(Inundated, levels = c("Yes", "No"))) %>%
  
ggplot(aes(x = Inundated, y = Total_biomass_g, fill = Soil_Treatment)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_color_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  scale_fill_manual(values = c("Organic" = "chartreuse4", "Sand" = "lightgoldenrod1")) +
  facet_wrap(~ Microbe_Treatment) +
  labs(
    title = "Total Biomass by Microbe Treatment and Inundation",
    x = "Inundation",
    y = "Total Biomass (g)",
    fill = "Soil Treatment"
  ) +
  theme_light()
```

# Generalized Linear Models

## Stem Height

```{r}
sp_data_final <- sp_data_final %>%
 filter(!is.na(Stem_Height_mm), 
         !is.na(Inundated), 
         !is.na(Soil_Treatment), 
         !is.na(Microbe_Treatment)) %>% 
  mutate(
    Inundated = as.factor(Inundated),
    Soil_Treatment = as.factor(Soil_Treatment),
    Microbe_Treatment = as.factor(Microbe_Treatment)
  )

# Fit GLM for Stem_Height_mm
stem_model <- glm(
  Stem_Height_mm ~ Inundated + Soil_Treatment + Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

# Display summary of the model
summary(stem_model)
```

## Belowground Biomass

```{r}
bg_model <- glm(
  Belowground_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

# Display summary of the model
summary(bg_model)
```

## Aboveground Biomass

```{r}
ag_model <- glm(
  Aboveground_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

# Display summary of the model
summary(ag_model)
```

## Total Biomass

```{r}
biomass_model <- glm(
  Total_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

# Display summary of the model
summary(biomass_model)
```

# Dredge 

## Stem Height

```{r}
sp_data_final <- na.omit(sp_data_final) # Removes rows with NAs

globalht_model <- glm(
  Stem_Height_mm ~ Inundated + Soil_Treatment + Microbe_Treatment + 
    Inundated:Soil_Treatment + Inundated:Microbe_Treatment + Soil_Treatment:Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

options(na.action = "na.fail")
model_set <- dredge(globalht_model)

print(model_set)
```

## Belowground Biomass

```{r}
sp_data_final <- na.omit(sp_data_final) # Removes rows with NAs

globalbm_model <- glm(
  Belowground_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment + 
    Inundated:Soil_Treatment + Inundated:Microbe_Treatment + Soil_Treatment:Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

options(na.action = "na.fail")
model_set <- dredge(globalbm_model)

print(model_set)
```

## Aboveground Biomass

```{r}
sp_data_final <- na.omit(sp_data_final) # Removes rows with NAs

globalbm_model <- glm(
  Aboveground_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment + 
    Inundated:Soil_Treatment + Inundated:Microbe_Treatment + Soil_Treatment:Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

options(na.action = "na.fail")
model_set <- dredge(globalbm_model)

print(model_set)
```

## Total Biomass

```{r}
sp_data_final <- na.omit(sp_data_final) # Removes rows with NAs

globalbm_model <- glm(
  Total_biomass_g ~ Inundated + Soil_Treatment + Microbe_Treatment + 
    Inundated:Soil_Treatment + Inundated:Microbe_Treatment + Soil_Treatment:Microbe_Treatment,
  family = gaussian(link = "identity"),
  data = sp_data_final
)

options(na.action = "na.fail")
model_set <- dredge(globalbm_model)

print(model_set)
```


```{r}
best_model <- get.models(model_set, subset = 1)[[1]]
summary(best_model) # Summary of the best model
```

