# BIOL 4800/6220 Fall 2024 R Tutorial

## Practical Computing Fall 2024

```{r}
setwd("C:/Users/michp/praccomp2024/")
```

### R Basics

#### Introduction

* R as a calculator

```{r}
1+100
```

```{r}
2^3
```

```{r}
7/2
```

```{r}
4-1
```

```{r}
2*3
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e3
```

```{r}
2/10000
```

```{r}
log(1)
```

```{r}
exp(1)
```

* Boolean Conditionals

```{r}
1 == 1
```

```{r}
2 == 1
```

```{r}
2 > 1
```

```{r}
2 <= 1
```

```{r}
 1 != 2
```

* Variables

```{r}
x <- 10
```

```{r}
x+x
```

```{r}
x - 5
```

```{r}
y <- x^x
```

```{r}
y
```

* Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- (1:5)
```

```{r}
2^v
```

```{r}
log(v)
```

* Environment

```{r}
ls()
```

```{r}
rm(v)
```

```{r}
ls()
```

#### Package Management

```{r}
#installed.packages()
```

```{r}
#install.packages("vegan", dependencies = TRUE)
library(vegan)
```


```{r}

```

#### Getting Help

**Search the Web**

```{r}
?plot
```
```{r}
help(plot)
```

```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```

#### Data and their formats

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5, 3.2),
                   likes.string = c(TRUE, FALSE, TRUE))
```

```{r}
cats
```

```{r}
getwd()
```

```{r}
write.csv(cats, "cats.csv")
```

```{r}
cats <- read.csv("cats.csv", header = TRUE, row.names = 1)
```

```{r}
cats
```

```{r}
cats$weight
```

```{r}
cats[, 2]
```

```{r}
cats[,"weight"]
```

```{r}
cats[c(2:3,1)]
```

```{r}
cats[c(1,3),c(2:3,1)]
```
```{r}
log(cats$weight)
```


```{r}
cats.log.weight <- log(cats$weight)
```

```{r}
cbind(cats,cats.log.weight)
```

```{r}
cats2 <- cbind(cats,cats.log.weight)
cats2
```


```{r}
paste("my cat is ", cats$coat, ", and it weighs ", cats$weight, " kg.", sep = "")
```

##### Data Types

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
typeof(cats$coat)
typeof(cats$weight)
typeof(cats$likes.string)
```

```{r}
typeof(1i)
```

```{r}
likes.pop <- c(1,0,1,1,0)
typeof(likes.pop)
```

```{r}
as.logical(likes.pop)
```

```{r}
names <- c("Iggy", "Bitey", "Garfield")
cats$names <- names
cats
```

```{r}
(names <- c(names, "Spike"))
cats
```

```{r}
(names2 <- c("Fluffy", names))
```

```{r}
1:10
```

```{r}
(z <- seq(20, 150, by =0.5))
```

```{r}
length(z)
```

```{r}
length(cats)
```

```{r}
length(cats$coat)
```

```{r}
dim(cats)
```

```{r}
typeof(z)
```

```{r}
class(z)
```

```{r}
head(z, n=20)
```

```{r}
tail(z, n=40)
```

```{r}
sort(z)
```

```{r}
sort(z, decreasing = TRUE)
```

```{r}
unique(z)
```

##### Data Frames

```{r}
cats
```

```{r}
cats$weight

typeof(cats$weight)
```

```{r}
str(cats$weight)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
```

```{r}
factor(coats)
```
```{r}
(categories <- factor(coats))
```

```{r}
str(categories)
```

##### Lists

```{r}
(list.example <- list(title="Numbers", numbers = seq(1,20, by=5),
                      data = TRUE))
```

```{r}
list.example$title
```

```{r}
data.frame(list.example)
```

##### Matrices
```{r}
(matrix.example <- matrix(0, nrow=3, ncol=5))
```

```{r}
class(matrix.example)
```

```{r}
typeof(matrix.example)
```

```{r}
dim(matrix.example)
```

```{r}
nrow(matrix.example)
ncol(matrix.example)
```

```{r}
data.frame(matrix.example)
```

##### Subsetting

```{r}
(p <- c(2.3, 6.9, 4.0, 23, 1))
```

```{r}
names(p) <- c("a", "b", "c", "d", "e")
p
```

```{r}
typeof(p)
class(p)
```

```{r}
p[1]
```

```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
(q <- p[-3])
```

```{r}
p[-(2:4)]
```

```{r}
p[c("a","e")]
```

```{r}
p[c(T,F,T,F,T)]
```

##### Factors

```{r}
(f <- factor(c("a", "b", "c", "d", "e")))
```

```{r}
f[f == "a"]
```

```{r}
f[1:3]
```

```{r}
(f2 <- factor(c("a", "a", "b", "c", "c")))
```

```{r}
f2[f2 == "a"]
```

```{r}
f2[f2 %in% c("a", "c")]
```

##### Matrices Revisited

```{r}
set.seed(4)
(m <- matrix(rnorm(6, 4, 0.3), nrow=6, ncol=4))
```

```{r}
hist(rnorm(10000, 4, 0.3))
```

```{r}
m[3:4, c(3,1)]
```

##### Lists Revisited

```{r}
(xlist <- list(a="BIOL48006220", b=rnorm(16,75,10), data="Grade"))
```

##### Data Frames Revisited

```{r}
(gp <- read.csv("data/forest_area_sq_km.csv"))
```

```{r}
head(gp, n = 10)
```

```{r}
head(gp[4], n = 3)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp[c(1,3)]
```

```{r}
gp[c(1,3), c(1,5:10)]
```

```{r}
gp[which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```


#### Conditionals and Flow

```{r}
n <- 10
n

if (n < 10) {
  print("n is less than 10.")
} else if (n > 10) {
  print("n is greater than 10.")
} else if (n == 10) {
  print("n is equal to 10.")
}
```

```{r}
for (n in seq(0,20)){
  if (n < 10) {
    print("n is less than 10.")
  } else if (n > 10) {
  print("n is greater than 10.")
  } else if (n == 10) {
  print("n is equal to 10.")
  }
}
```

```{r}
for (n in rnorm(20,10,3)){
  if (n < 10) {
    print(paste0(n, "n is less than 10."))
  } else if (n > 10) {
  print(paste0(n, "n is greater than 10."))
  } else if (n == 10) {
  print(paste0(n, "n is equal to 10."))
  }
}
```

```{r}
g <- 0
while (g <= 10) {
  print(paste0(g, " is less than or equal to 10.", sep=" "))
  g <- g+1
}
```

#### Plotting and Figures

```{r}
gap <- read.csv("data/gapminder_all.csv")
```
 
```{r}
gap
```
 
```{r}
library(ggplot2)
```
 
```{r}
ggplot(data=gap, mapping = aes(y=gdpPercap_1982, x=pop_1982)) +
  geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv")

gapminder
```

```{r}
ggplot(data=gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) +
  geom_point(aes(color=continent))
```

```{r}
ggplot(data=gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) +
  geom_point(aes(color=continent), alpha = 0.25) + 
  scale_x_log10() + geom_smooth(method = lm, color="black", fill="grey") +
  theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by = country)) +
  geom_line(mapping=aes(color=continent)) +
  geom_point() +
  theme_minimal()
```

```{r}
library(ggplot2)
ggplot(data=gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) +
  geom_point(color="darkblue", alpha = 0.25, size =1) + 
  scale_x_log10() + geom_smooth(method = lm, color="deeppink", fill="chartreuse", size = 0.5) +
  theme_classic()
```

```{r}
library(ggplot2)
ggplot(data=gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) +
  geom_point(color="darkblue", alpha = 0.25, size =1) + 
  geom_smooth(method = loess, color="deeppink", fill="chartreuse", size = 0.5) +
  theme_classic()
```

```{r}
unique(gapminder$continent)
```

```{r}
asia <- gapminder[gapminder$continent == "Asia", ]
```

```{r}
ggplot(data = asia, mapping = aes(x=year, y=lifeExp)) +
  labs(
    x = "Year", 
    y = "Life Expectancy (years)",
    title = "Life Expectancy over time in Asian countries"
  ) +
  geom_line(color="purple") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
ggsave("data/AsiaLifeExp.png")
```

```{r}
write.csv(asia, file = "data/asia.csv")
write.table(gapminder, file="data/gapminder_web_20240930.csv", sep= ",")
```

###### Fancy Plots

```{r}
install.packages(c("ggridges", "viridis", "hrbrthemes"), dependencies = TRUE)
```

```{r}
library(ggridges)
library(viridis)
library(ggplot2)
```

```{r}
ggplot(lincoln_weather, aes(x=`Mean Temperature [F]`, y = `Month`, fill = stat(x))) +
  labs(title = "Temperature in Lincoln, NE in 2016") +
  geom_density_ridges_gradient(scales = 3, rel_min_height=0.01) +
  scale_fill_viridis_c(name = "Temp. [F]", option = "C")
```

```{r}
install.packages(c("ggstatsplot", "tidyverse", "palmerpenguins"))
```

```{r}
data("penguins", package = "palmerpenguins")
penguins
```

```{r}
library(tidyverse)
(penguins2 <- drop_na(penguins))
```

```{r}
ggplot(penguins, aes(x=`bill_length_mm`, y=`bill_depth_mm`)) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r}
install.packages("rlang", dependencies = TRUE)
install.packages("ggstatsplot", dependencies = TRUE)
library(ggstatsplot)
plt <- ggbetweenstats(
  data = penguins2,
  x = species,
  y = bill_length_mm
)
plt

```

```{r}
library(ggridges)
library(ggplot2)
library(viridis)
spider_data <- read.csv("data/spider_data_20221101.csv")
spider_data
```

```{r}
library(tidyverse)
spider_data %>% 
  arrange(year) %>% 
  mutate(name = factor(family, levels = unique(family))) %>% 
  ggplot(aes(x = `year`, y = `family`, color = `family`, fill = 
                          `family`)) + geom_density_ridges(alpha = .5, stat = "binline", binwidth = 25) +
    theme(legend.position = "none",
        strip.text.x = element_text(size=5)
        )
```

```{r}
(penguins3 <- with(penguins2, cbind(bill_length_mm, bill_depth_mm, 
                                   flipper_length_mm, body_mass_g)))
head(penguins3)
penguin_pca <- princomp(penguins3)
summary(penguin_pca)
```

```{r}
loadings(penguin_pca)
```

```{r}
biplot(penguin_pca)
```

```{r}
#install.packages("devtools")
library(devtools)
install_github("vqv/ggbiplot")
library(ggbiplot)
ggbiplot(penguin_pca,
         groups = penguins2$species, ellipse = TRUE, scale. = TRUE) +
  scale_color_discrete(name = "")
```

```{r}
#install.packages("gganimate")
#install.packages("gifski")

# Get data:
library(gapminder)
 
# Charge libraries:
library(ggplot2)
library(gganimate)
 
# Make a ggplot, but add frame=year: one image per year
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10() +
  theme_bw() +
  # gganimate specific bits:
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')

# Save at gif:
anim_save("271-ggplot2-animated-gif-chart-with-gganimate1.gif")
```

### Statistics

```{r}
rand <- rnorm(50, 30, 8)
rand
```
```{r}
sum(rand)
```

```{r}
sum(rand)/length(rand)
mean(rand)
```
```{r}
sort(rand)
```
```{r}
min(rand)
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
sqrt(rand)
```

```{r}
log(rand)
```

```{r}
#things <- scan()
things
```

#### Analyses

##### Univariate Statistics

1. Categorical Data
  - Barplots
  
```{r}
beer <- c(3,4,1,1,3,4,3,3,1,3,2,1,2,1,2,3,2,3,1,1,1,1,4,3,1)
length(beer)
table(beer)
```
  
```{r}
barplot(table(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
pie(table(beer), main = "Beer preference by students")
```

2. Numerical Data

```{r}
stem(rand)
```

```{r}
stripchart(rand)
```

3. Measures of center

```{r}
mean(rand)
```

```{r}
median(rand)
```

```{r}
which(table(rand) == max(table(rand)))
```

4. Variation
  - Range

```{r}
range(rand)
```

  - Variance

```{r}
var(rand)
```

```{r}
sd(rand)
```

  - IQR

```{r}
IQR(rand)
```

  - z-score

```{r}
scale(rand)
```
  
  - Summary
  
```{r}
summary(rand)
```

5. Plots
  - Histograms
  
```{r}
hist(rand, breaks = 15)
```

  - Box plots
  
```{r}
boxplot(rand)
```

```{r}
#install.packages("vioplot")
library(vioplot)
vioplot(rand, col = "maroon4")
```

```{r}
vioplot(gapminder$lifeExp~gapminder$continent, col = "purple")
```

#### Bivariate and multivariate statistics

  - Plotting and regression
    - Box plotting

```{r}
boxplot(gapminder$lifeExp~gapminder$continent, col = "darkorchid4")
```

```{r}
(spid.gen <- read.csv("data/spider_genitalia.csv", header = TRUE))
```

```{r}
vioplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat, col = "darkred")
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, col = "royalblue3")
bulb.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
abline(bulb.reg, col = "firebrick1")
summary(bulb.reg)
```

```{r}
sqrt(0.5115)
```

```{r}
sqrt(bulb.reg$coefficients)
```

```{r}
cor(spid.gen$left.bulb, spid.gen$right.bulb, method = "spearman")
```

```{r}
t.test(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

```{r}
summary(rand)
t.test(rand, mu = 30)
```

```{r}
wilcox.test(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

```{r}
shapiro.test(rand)
```

```{r}
summary(lm(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat))
```

```{r}
habitat <- c("A", "A", "A", "A", "A", "B", "B", "B", "B", "B", "C", "C", "C", "C", "C")

parload <- c(10, 12, 8, 11, 14, 14, 16, 10, 9, 12, 18, 19, 20, 21, 22)

parhab <- cbind(habitat, parload)

parhab <- data.frame(parhab)

parhab$habitat <- as.factor(parhab$habitat)
parhab$parload <- as.numeric(parhab$parload)

parhab
```

```{r}
parhab.aov <- aov(parhab$parload ~ parhab$habitat)
summary(parhab.aov)
```

```{r}
(parhab.tukey <- TukeyHSD(parhab.aov))
```

```{r}
#install.packages("glue")
library(ggstatsplot)
ggbetweenstats(parhab, x = habitat, y = parload)
```

```{r}
pvals <- c(0.05, 0.005, 0.0005, 0.00005, 0.000005)

p.adjust(pvals, method ="bonferroni")

p.adjust(pvals, method = "fdr")
```

```{r}
gapminder.2007 <- gapminder[which(gapminder$year == 2007),]
gapminder.2007.pop.gdpPercap.lifeExp.continent <- lm(gapminder.2007$pop ~ gapminder.2007$gdpPercap + gapminder.2007$lifeExp + gapminder.2007$continent)

summary(gapminder.2007.pop.gdpPercap.lifeExp.continent)
```

```{r}
#install.packages("MuMIn", dependencies = TRUE)
library(MuMIn)
# prevent fitting sub-models to different datasets
oop <- options(na.action = "na.fail")

dd <- dredge(gapminder.2007.pop.gdpPercap.lifeExp.continent)
subset(dd, delta < 2)
```

```{r}
#visualize the model selection table:

if(require(graphics)) {
  par(mar = c(3,5,6,4))
  plot(dd, labAsExpr = TRUE)
}
```

```{r}
# Model average models with delta AICc < 2
model.avg(dd, subset = delta < 2)

#or as a 95% confidence set
model.avg(dd, subset = cumsum(weight) <= .95) #get averaged coefficient

#'Best' model
summary(get.models(dd, 1) [[1]])

```

### Geospatial Mapping

```{r}
install.packages("maps", dependencies = TRUE)
library(maps)

install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata", "rgbif", "terra", "geodata", "sdmpredictors", "stringr", "stars", "elevatr", "tigris", "data.table"))

devtools::install_github("yutannihilation/ggsflabel")

library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgbif)
library(terra)
library(geodata)
library(sdmpredictors)
library(stringr)
library(raster)
library(ggspatial)
library(stars)
library(elevatr)
library(tigris)
library(ggsflabel)
library(data.table)
```

```{r}
geom_relief <- function(mapping = NULL, data = NULL,
                        stat = "identity", position = "identity",
                        ...,
                        raster = TRUE,
                        interpolate = TRUE,
                        na.rm = FALSE,
                        show.legend = NA,
                        inherit.aes = TRUE) {
   ggplot2::layer(
      data = data,
      mapping = mapping,
      stat = stat,
      geom = GeomRelief,
      position = position,
      show.legend = show.legend,
      inherit.aes = inherit.aes,
      params = list(
         raster = raster,
         interpolate = interpolate,
         na.rm = na.rm,
         ...
      )
   )
}

GeomRelief <- ggplot2::ggproto("GeomRelief", GeomTile,
  required_aes = c("x", "y", "z"),
  default_aes = ggplot2::aes(color = NA, fill = "grey35", size = 0.5, linetype = 1,
                             alpha = NA, light = "white", dark = "gray20", sun.angle = 60),
  draw_panel = function(data, panel_scales, coord, raster, interpolate) {
     if (!coord$is_linear()) {
        stop("non lineal coordinates are not implemented in GeomRelief", call. = FALSE)
     } else {
        coords <- as.data.table(coord$transform(data, panel_scales))
        
        # This is the only part that's actually new. The rest is essentially 
        # copy-pasted from geom_raster and geom_tile
        coords[, sun.angle := (sun.angle + 90)*pi/180]
        coords[, dx := .derv(z, x), by = y]
        coords[, dy := .derv(z, y), by = x]
        coords[, shade := (cos(atan2(-dy, -dx) - sun.angle) + 1)/2]
        coords[is.na(shade), shade := 0]
        coords[, fill := .rgb2hex(colorRamp(c(dark, light), space = "Lab")(shade)),
               by = .(dark, light)]
        
        # From geom_raster and geom_tile
        if (raster == TRUE){
           if (!inherits(coord, "CoordCartesian")) {
              stop("geom_raster only works with Cartesian coordinates", call. = FALSE)
           }
           # Convert vector of data to raster
           x_pos <- as.integer((coords$x - min(coords$x)) / resolution(coords$x, FALSE))
           y_pos <- as.integer((coords$y - min(coords$y)) / resolution(coords$y, FALSE))
           
           nrow <- max(y_pos) + 1
           ncol <- max(x_pos) + 1
           
           raster <- matrix(NA_character_, nrow = nrow, ncol = ncol)
           raster[cbind(nrow - y_pos, x_pos + 1)] <- alpha(coords$fill, coords$alpha)
           
           # Figure out dimensions of raster on plot
           x_rng <- c(min(coords$xmin, na.rm = TRUE), max(coords$xmax, na.rm = TRUE))
           y_rng <- c(min(coords$ymin, na.rm = TRUE), max(coords$ymax, na.rm = TRUE))
           
           grid::rasterGrob(raster,
                            x = mean(x_rng), y = mean(y_rng),
                            width = diff(x_rng), height = diff(y_rng),
                            default.units = "native", interpolate = interpolate
           )
           
        } else {
           ggplot2:::ggname("geom_rect", grid::rectGrob(
              coords$xmin, coords$ymax,
              width = coords$xmax - coords$xmin,
              height = coords$ymax - coords$ymin,
              default.units = "native",
              just = c("left", "top"),
              gp = grid::gpar(
                 col = coords$fill,
                 fill = alpha(coords$fill, coords$alpha),
                 lwd = coords$size * .pt,
                 lty = coords$linetype,
                 lineend = "butt"
              )
           ))
           
        }
     }
  }
)

rect_to_poly <- function(xmin, xmax, ymin, ymax) {
   data.frame(
      y = c(ymax, ymax, ymin, ymin, ymax),
      x = c(xmin, xmax, xmax, xmin, xmin)
   )
}

.rgb2hex <- function(array) {
   rgb(array[, 1], array[, 2], array[, 3], maxColorValue = 255)
}


.derv <- function(x, y, order = 1, cyclical = FALSE, fill = FALSE) {
   N <- length(x)
   d <- y[2] - y[1]
   if (order >= 3) {
      dxdy <- .derv(.derv(x, y, order = 2, cyclical = cyclical, fill = fill),
                    y, order = order - 2, cyclical = cyclical, fill = fill)
   } else {
      if (order == 1) {
         dxdy <- (x[c(2:N, 1)] - x[c(N, 1:(N-1))])/(2*d)
      } else if (order == 2) {
         dxdy <- (x[c(2:N, 1)] + x[c(N, 1:(N-1))] - 2*x)/d^2
      }
      if (!cyclical) {
         if (!fill) {
            dxdy[c(1, N)] <- NA
         }
         if (fill) {
            dxdy[1] <- (-11/6*x[1] + 3*x[2] - 3/2*x[3] + 1/3*x[4])/d
            dxdy[N] <- (11/6*x[N] - 3*x[N-1] + 3/2*x[N-2] - 1/3*x[N-3])/d
         }
      }
      
   }
   return(dxdy)
}
```

```{r}
theme_set(theme_bw())

world <- ne_countries(scale = "medium", returnclass = "sf")

#water <- st_read("data/USA_Detailed_Water_Bodies.shp")

world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
  geom_sf() +
  #geom_sf(data = water, fill = "lightblue") +
  geom_text(data= world_points,
            aes(x=X, y=Y, label=name),
            color = "forestgreen", 
            fontface = "bold", 
            size = 2,
            check_overlap = TRUE)
```

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

world_relief <- get_elev_raster(locations = world, z = 1, clip = "locations")

#r <- raster(xmn = -105, xmx = -50, ymn = 0, ymx = 55, nrow = 100, ncol = 100)
#r[]=runif(100*100)

plot(world_relief)
plot(ne_coastline(), add=TRUE)
#plot(r, add=TRUE)
```

```{r}
nc_counties <- counties(state = 'North Carolina', cb = FALSE, progress_bar = TRUE)
nc_counties
```

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

# Use elevatr to get elevation data
nc_elevations <- elevatr::get_elev_raster(
  locations = nc_counties, z = 8, clip = 'locations'
)

plot(nc_elevations)

# Convert elevation data to dataframe
nc_elevations <- as.data.frame(nc_elevations, xy = TRUE)
colnames(nc_elevations)[3] <- 'elevation'

# Remove rows with one or more NA's using complete.cases
nc_elevations <- nc_elevations[complete.cases(nc_elevations), ]
```

```{r}
ggplot() +
  geom_relief(data = nc_elevations,
              aes(x = x, y = y, z = elevation, light = 'white', dark = 'grey20'), 
               raster = FALSE, interpolate = TRUE, sun.angle = 60) +
  #geom_sf(data = water, color = NA, fill = 'lightblue', size = 0.6, alpha = 1) +
  geom_sf(data = nc_counties, color = 'white', fill = NA) +
  geom_sf_text_repel(data = nc_counties, aes(label = NAME),
                     size = 3, bg.color = 'white', color = 'darkgreen') +
  coord_sf(datum = sf::st_crs(nc_counties), ) +
  labs(title = 'Land Elevations',
       subtitle = 'Counties of North Carolina',
       caption = 'Data source: https://registry.opendata.aws/terrain-tiles')
```

```{r}
Spartina.alterniflora_GBIF <- occ_data(scientificName = "Spartina alterniflora", hasCoordinate = TRUE, decimalLongitude = "-100, -60", decimalLatitude = "0, 50", limit = 10000)

Spartina.alterniflora_GBIF <- data.frame(Spartina.alterniflora_GBIF$data)
```

```{r}
states <- states()

zone <- states[ which(states$STUSPS == "NC" | states$STUSPS == "VA" | states$STUSPS == "MD"),]

zone <- st_make_valid(zone)
```

```{r}
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet))

# Use elevatr to get elevation data
elevations <- elevatr::get_elev_raster(
  locations = zone, z = 8, clip = 'locations'
)

# Convert elevation data to dataframe
elevations <- as.data.frame(elevations, xy = TRUE)
colnames(elevations)[3] <- 'elevation'

# Remove rows with one or more NA's using complete.cases
elevations <- elevations[complete.cases(elevations), ]
```

```{r}
ggplot() +
  geom_relief(data = elevations,
              aes(x = x, y = y, z = elevation, light = 'white', dark = 'grey20'), 
               raster = FALSE, interpolate = TRUE, sun.angle = 60) +
  #geom_sf(data = water, color = NA, fill = 'lightblue', size = 0.6, alpha = 1) +
  geom_sf(data = zone, color = 'white', fill = NA) +
  geom_sf_text_repel(data = zone, aes(label = NAME),
                     size = 3, bg.color = 'white', color = 'skyblue') +
  coord_sf(datum = sf::st_crs(zone)) +
  geom_point(data = Spartina.alterniflora_GBIF, aes(x = decimalLongitude, y = decimalLatitude, color='green')) +
  labs(title = 'Land Elevations',
       subtitle = 'State in Zone of Interest',
       caption = 'Data source: https://registry.opendata.aws/terrain-tiles')
```

# ENM workbook

## Citation
- GBIF.org (12 October 2023) GBIF Occurrence Download  https://doi.org/10.15468/dl.3stbgp

## Download, read, and explore occurence data

```{r}
Spartina.alterniflora_GBIF <- occ_search(scientificName = "Spartina alterniflora", hasCoordinate = TRUE, decimalLongitude = "-100, -60", decimalLatitude = "0, 50", limit = 10000)

write.csv(Spartina.alterniflora_GBIF$data, file = "Spartina.alterniflora_GBIF.csv", row.names = FALSE)

Spartina.alterniflora_GBIF_df <- read.csv("Spartina.alterniflora_GBIF.csv")

#PROW_GBIF_df <- read.csv("0011132-231002084531237/0011132-231002084531237.csv")

Spartina.alterniflora_GBIF_df
```

```{r}
colnames(Spartina.alterniflora_GBIF_df)
```

### Construct SDM

#### Obtain Environmental Layers
a. View available layers

```{r}
# Inspect the available datasets and layers
datasets <- list_datasets(terrestrial = TRUE, marine = FALSE)
View(datasets)
layers <- list_layers(datasets)
View(layers)
```
b. Choose and load layers

```{r}
# Load equal area rasters
layercodes <- c("WC_alt", "WC_bio1", "WC_bio2", "WC_bio3", "WC_bio4", "WC_bio5", "WC_bio6", "WC_bio7", "WC_bio8", "WC_bio9", "WC_bio10", "WC_bio11", "WC_bio12", "WC_bio13", "WC_bio14", "WC_bio15", "WC_bio16", "WC_bio17", "WC_bio18", "WC_bio19")
env <- load_layers(layercodes,  datadir = "sdm_data_layers")
#plot(env)

easternUS <- raster::crop(env, raster::extent(-100, -60, 0, 50))
plot(easternUS)
```

c.
```{r}
# Compare correlations between predictors for Spartina
prettynames <- list(WC_alt="Altitude", 
                    WC_bio1="Annual mean temp", 
                    WC_bio2="Mean diurnal temperature range", 
                    WC_bio3="Isothermality", 
                    WC_bio4="Temperature seasonality", 
                    WC_bio5="Maximum temperature", 
                    WC_bio6="Minimum temperature", 
                    WC_bio7="Annual temperature range", 
                    WC_bio8="Mean temperature of wettest quarter", 
                    WC_bio9="Mean temperature of driest quarter", 
                    WC_bio10="Mean temperature of warmest quarter", 
                    WC_bio11="Mean temperature of coldest quarter", 
                    WC_bio12="Annual precipitation", 
                    WC_bio13="Precipitation of wettest month", 
                    WC_bio14="Precipitation of driest month", 
                    WC_bio15="Precipitation seasonality", 
                    WC_bio16="Precipitation of wettest quarter", 
                    WC_bio17="Precipitation of driest quarter", 
                    WC_bio18="Precipitation of warmest quarter", 
                    WC_bio19="Precipitation of coldest quarter")

easternUS_correlations <- pearson_correlation_matrix(easternUS)

p1 <- plot_correlation(easternUS_correlations, prettynames)

cowplot::plot_grid(p1, labels = "Eastern US", ncol = 1, nrow = 1)

print(correlation_groups(easternUS_correlations))
```

#### Choose occurence points for SDMs
```{r}
summary(Spartina.alterniflora_GBIF_df)
```

```{r}
obs_data <- Spartina.alterniflora_GBIF_df[!is.na(PROW_GBIF_df$decimalLatitude), ]
```

#### Crop environmental layers to fit occurences
```{r}
# Determine geographic extent of our data
max_lat <- ceiling(max(obs_data$decimalLatitude))
min_lat <- floor(min(obs_data$decimalLatitude))
max_lon <- ceiling(max(obs_data$decimalLongitude))
min_lon <- floor(min(obs_data$decimalLongitude))
# Store boundaries in a single extent object
geographic_extent <- raster::extent(x = c(min_lon, max_lon, min_lat, max_lat))

sample_extent <- geographic_extent * 1.25

sample_extent
```

```{r}
# Fetch occurrences and prepare for ???
points <- cbind(obs_data$decimalLongitude, obs_data$decimalLatitude)
#points <- spTransform(points)
occfile <- tempfile(fileext = ".csv")
write.csv(cbind(points, value=1), occfile)
#read.csv("occfile")
```

```{r}
# Download data with geodata's world function to use for our base map
world_map <- world(resolution = 3,
                   path = "data/")

# Crop the map to our area of interest
my_map <- crop(x = world_map, y = sample_extent)

# Plot the base map
plot(my_map,
     axes = TRUE, 
     col = 'grey95')

# Add the points for individual observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = 'springgreen4', 
       pch = 20, 
       cex = 0.75)
```

#### FIX!!!

```{r}
bioclim_data <- worldclim_global(var = "bio",
                                 res = 2.5,
                                 path = "0011132-231002084531237/")
```

```{r}
# Make an extent that is 25% larger
sample_extent <- geographic_extent * 1.25

# Crop bioclim data to desired extent
bioclim_data <- crop(x = env, y = sample_extent)

# Plot the first of the bioclim variables to check on cropping
plot(bioclim_data[[1]])
```

```{r}
bioclim_data_SR <- as(bioclim_data, "SpatRaster")
```

```{r}
# Set the seed for the random-number generator to ensure results are similar
set.seed(20210707)

# Randomly sample points (same number as our observed points)
background <- spatSample(x = bioclim_data_SR,
                         size = 100000,    # generate 1,000 pseudo-absence points
                         values = FALSE, # don't need values
                         na.rm = TRUE,   # don't sample from ocean
                         xy = TRUE)      # just need coordinates

# Look at first few rows of background
head(background)
```
```{r}
# Plot the base map
plot(my_map,
     axes = TRUE, 
     col = "grey95")

# Add the background points
points(background,
       col = "grey30",
       pch = 2,
       cex = 0.1,)

# Add the points for individual observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = "firebrick", 
       pch = 20, 
       cex = 0.25)
```

```{r}
# Pull out coordinate columns, x (longitude) first, then y (latitude) from presences
presence <- obs_data[, c("decimalLongitude", "decimalLatitude")]
# Add column indicating presence
presence$pa <- 1

# Convert background data to a data frame
absence <- as.data.frame(background)
# Update column names so they match presence points
colnames(absence) <- c("decimalLongitude", "decimalLatitude")
# Add column indicating absence
absence$pa <- 0

# Join data into single data frame
all_points <- rbind(presence, absence)

#Convert column names to match bioclim
colnames(all_points) <- c("longitude", "latitude","pa")

# Reality check on data
head(all_points)
```

```{r}
bioclim_extract <- extract(x = bioclim_data_SR,
                           y = all_points[, c("longitude", 
                                              "latitude")],
                           ID = FALSE)
```

```{r}
# Add the point and climate datasets together
points_climate <- cbind(all_points, bioclim_extract)

# Identify columns that are latitude & longitude
drop_cols <- which(colnames(points_climate) %in% c("longitude", "latitude"))
drop_cols # print the values as a reality check
```

```{r}
# Remove the geographic coordinates from the data frame
points_climate <- points_climate[, -drop_cols]
```

```{r}
# Create vector indicating fold
install.packages("predicts", dependencies = T)
library(predicts)

fold <- folds(x = points_climate,
              k = 5,
              by = points_climate$pa)

table(fold)
```

```{r}
testing <- points_climate[fold == 1, ]
training <- points_climate[fold != 1, ]
```

```{r}
# Build a model using training data
glm_model <- glm(pa ~ ., data = training, family = binomial())
```

```{r}
# Get predicted values from the model
glm_predict <- predict(bioclim_data_SR, glm_model, type = "response")

# Print predicted values
plot(glm_predict)
```

```{r}
# Use testing data for model evaluation
glm_eval <- pa_evaluate(p = testing[testing$pa == 1, ],
                        a = testing[testing$pa == 0, ],
                        model = glm_model,
                        type = "response")
```

```{r}
# Determine minimum threshold for "presence"
glm_threshold <- glm_eval@thresholds$max_spec_sens
```

```{r}
# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = "springgreen4")

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")
```

```{r}
glm_predict > glm_threshold
```

```{r}
# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "springgreen4")) # <-- Update the values HERE

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       #col = alpha("blue",0.5),
       col = "blue",
       pch = ".", 
       cex = 0.0000000000001
       )

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")
```
```{r}
# Save PDF
pdf("Spartina.alterniflora_GLM_Pred.pdf")

# Plot base map
plot(my_map, 
     axes = TRUE, 
     col = "grey95")

# Only plot areas where probability of occurrence is greater than the threshold
plot(glm_predict > glm_threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "springgreen4")) # <-- Update the values HERE

# And add those observations
points(x = obs_data$decimalLongitude, 
       y = obs_data$decimalLatitude, 
       #col = alpha("blue",0.5),
       col = "blue",
       pch = ".", 
       cex = 0.0000000000001
       )

# Redraw those country borders
plot(my_map, add = TRUE, border = "grey5")

dev.off()
```
### Tidyverse

```{r}
library(tidyverse)
library(palmerpenguins)
head(penguins)
```

```{r}
weather <- read.csv("http://pasta.lternet.edu/package/data/eml/knb-lter-pal/28/8/375b34051b162d84516ec2d02f864675")
```

- Structure

```{r}
str(penguins)
```

```{r}
str(weather)
```

```{r}
colnames(penguins)
```

```{r}
dim(penguins)
```

```{r}
penguins %>% select(year)
```

```{r}
penguins %>% select(-year)

```

```{r}
penguins %>% select(species, island, year) %>% filter(year == 2007)
```

```{r}
penguins %>% select(species, island, year) %>% filter(year == 2007 | year == 2009)
```

```{r}
torg_2007_penguins <- penguins %>% select(species, island, year) %>% filter(year == 2007 & island == "Torgersen")
```

```{r}
is.na(penguins$sex)
```

```{r}
penguins %>% filter(is.na(sex))
```

```{r}
penguins %>% filter(!is.na(sex) & body_mass_g >= 5000)
```

```{r}
summary(penguins %>% filter(!is.na(sex) & body_mass_g >= 5000))
```

```{r}
penguins %>% mutate(bill.mass.ratio = bill_length_mm/body_mass_g)
```

```{r}
penguins_bill.mass.ratio <- penguins %>% mutate(bill.mass.ratio = bill_length_mm/body_mass_g)
vioplot(penguins_bill.mass.ratio$bill.mass.ratio)
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>% 
  mutate(body_size_cat = case_when(body_mass_g >= 5000 ~ "big",
                                              body_mass_g <= 3500 ~ "small",
                                              body_mass_g < 5000 & body_mass_g > 3500 ~ "medium",
                                              TRUE ~ "NA"))
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>% 
  mutate(body_size_cat = case_when(body_mass_g >= 5000 ~ "big",
                                              body_mass_g <= 3500 ~ "small",
                                              body_mass_g < 5000 & body_mass_g > 3500 ~ "medium",
                                              TRUE ~ "NA")) %>%
  filter(!is.na(sex)) %>% 
  group_by(species, sex) %>% 
  summarize(mean_body_mass_g = mean(body_mass_g, na.rm = TRUE), sd_body_mass_g = sd(body_mass_g))
```

- Manipulating Data

```{r}
str(weather)
```

```{r}
year("2024-10-28")
ymd("2024-10-28")
yday("2024-10-28")
```

```{r}
weather %>% mutate(Date = yday(Date)) %>% group_by(Date) %>% summarize(mean(Temperature.Average..C.))
```

```{r}
weather %>%  mutate(Date =ymd(Date)) %>% 
  mutate(Date - floor_date(Date, unit = 'year')) %>%
  group_by(Date) %>% 
  summarize(mean(Temperature.Average..C.)) %>%

ggplot(aes(x=Date, y=`mean(Temperature.Average..C.)`)) +
  geom_point(aes(color = year(Date))) +
  scale_color_viridis_c() +
  geom_line(color = "white") +
  geom_smooth(method="lm", linewidth = 1, linetype = 1, color = "firebrick")
```

```{r}
colnames(penguins)
```

```{r}
colnames(weather)
```

```{r}
(weather.sums <- weather %>% 
  group_by(year(Date)) %>%
  summarize(avg.temp = mean(Temperature.Average..C., na.rm = TRUE),
                   max.temp = mean(Temperature.High..C., na.rm = TRUE),
                   min.temp = mean(Temperature.Low..C., na.rm = TRUE)) %>%
  rename(year = `year(Date)`)
)
```

```{r}
(penguins_weather.sums <- left_join(penguins, weather.sums, by = "year"))
```

```{r}
penguins_weather.sums %>%
  filter(!is.na(sex)) %>% 
  
ggplot(aes(x = sex, y = body_mass_g, color = sex, fill = sex)) +
  geom_violin() + 
  #scale_color_viridis_d()
  scale_color_manual(values = c("darkorange", "darkviolet")) +
  scale_fill_manual(values = c("darkorange", "darkviolet")) +
  facet_wrap(~species + year) +
  labs(title = "Average penguin body size per species per year by sex", 
       x = "Sex", y = "Body Mass (g)")
  
```

```{r}
penguins_weather.sums %>%
  filter(!is.na(sex)) %>%
  group_by(species, sex, year) %>%
  summarize(mean.body.mass.g = mean(body_mass_g), mean.min.temp.C = mean(min.temp)) %>% 
  ggplot(aes(x = mean.min.temp.C, y = mean.body.mass.g, color = sex, fill = sex)) +
  geom_point() +
  scale_color_manual(values = c("darkorange", "darkviolet")) +
  scale_fill_manual(values = c("darkorange", "darkviolet")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~species) +
  theme_classic() +
  labs(title = "Average penguin body mass by average annual minimum temperature",
       x="Average minimum temperature",
       y="Average body mass")
  
```

